Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices

### 2022-10-18 supp/index.Rmd
<edit index>

Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast. Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area. Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material. The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class. Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This suggests that a shared, and consistent, mental template was maintained independent of biface size. The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does. This finding supports the argument that elliptical biface morphology diverges through local reduction practices. As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.

@@

American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking

@@

@@

Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.2.1 [@R] using the `geomorph` library v. 4.0.4 [@RN11530;@RN1774;@RN9565]. Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space following alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].

@@

Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces. The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference image or 3D mesh [@RN1731;@RN479].

>>>>

Elliptical bifaces are regularly found at Mississipian sites throughout the American Southeast. Those from Millsap Cache and Jowell Farm comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area. Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material. The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class. Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This suggests that a shared, and consistent, mental template was maintained independent of biface size. The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does. This finding supports the argument that elliptical biface morphology diverges through local reduction practices. As expected, greater shape diversity occurs in the small class, where small bifaces from Jowell Farm were found to be more morphologically diverse than small bifaces from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences that occur in elliptical bifaces.

@@

American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking, 41AN13

@@

The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \@ref(fig:figlm)). Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R [@R] and the `StereoMorph` package [@RN9091].

@@

Landmark data were aligned to a global coordinate system [@RN8477;@RN7502;@RN11622;@RN11623;@RN11563], achieved through generalised Procrustes superimposition [@RN11138;@RN478;@RN1646] in R using the `geomorph` and `RRPP` packages [@RN1655;@RN11775;@RN11530;@RN1774;@RN9565]. Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid [@RN11139;@RN11140;@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN11141;@RN11142;@RN1646;@RN11563].

@@

Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces, and the scatterplot represents the dispersion of shapes in tangent space [@RN8633;@RN5616;@RN11143;@RN7550]. Those shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference image [@RN1731;@RN479].
